---
title: "Proyecto M1"
---

## Setup
<!-- PROTIP: en vez de picarle al botón de Preview, usar "quarto preview <filename> --no-browser", sin el --no-watch-inputs -->

```{r}
library(fpp3)
mi_serie <- aus_retail |> filter(`Series ID` == "A3349397X")
print(mi_serie)
tail(mi_serie)
```

## Plot of turnover

```{r}
mi_serie |> autoplot(Turnover)
```

We can notice three things about the series:

1. An upward trend.
2. A peak every year. Reading the data, it seems to be on December. It might have seasonality.
3. The peaks become wider as time goes by. That means the series is heteroskedastic.

## Seasonal plot

```{r}
mi_serie |> gg_season(Turnover)
```

This plot confirms not only that our series has yearly seasonality with a peak on December, it also reveals a minor peak on June.

## Subseries

```{r}
mi_serie |> gg_subseries(Turnover)
```

This plot shows that December is the month with the highest average.

## Autocorrelation

```{r}
mi_serie |> ACF(Turnover, lag_max=48) |> autoplot()
```

## Result of EDA

- The series **has a trend** (increasing).
- The series **is seasonal yearly** (the autocorrelation function peaks every 12 time units [months]).
- The series **shows a minor seasonal peak on June**.
- The series is **heteroskedastic** (the seasonal peaks are increasingly sharper as time goes by).
