---
title: "Proyecto M1"
---

## About the series

<!-- Descripción completa del dataset. Pistas: (NSW household goods retail turnover), (millions of AUD), range (Apr 1982–Dec 2018), number of observations (441) -->

## Setup
<!-- PROTIP: en vez de picarle al botón de Preview, usar "quarto preview <filename> --no-browser", sin el --no-watch-inputs -->

```{r}
library(fpp3)
mi_serie <- aus_retail |> filter(`Series ID` == "A3349397X")
print(mi_serie)
tail(mi_serie)
```

## Plot of turnover

```{r}
mi_serie |> autoplot(Turnover) + labs(title="Household good retail turnover in New South Wales, Australia", y="Sales turnover")
```

We can notice three things about the series:

1. An upward trend.
2. A peak every year. Reading the data, it seems to be on December. It might have seasonality.
3. The amplitude of the peaks increases as time goes by. That means the series is heteroskedastic.

## Seasonal plot

```{r}
mi_serie |> gg_season(Turnover) + labs(title="Household good retail turnover in New South Wales, Australia (turnover per year)", y="Sales turnover each year")
```

This plot confirms not only that our series has yearly seasonality with a peak on December, it also reveals a minor peak on June.

## Subseries

```{r}
mi_serie |> gg_subseries(Turnover) + labs(title="Household good retail turnover in New South Wales, Australia, grouped by month", y="Sales turnover")
```

This plot shows that December is the month with the highest average. December is also the month that rises more steeply than the other months. That means the series is **heteroskedastic** (the seasonal effect scales with time, making the series's variance variable).

## Autocorrelation

```{r}
mi_serie |> ACF(Turnover, lag_max=48) |> autoplot() + labs(title="Autocorrelation of household good retail turnover in New South Wales, Australia", y="Value of autocorrelation function")
```

We can see in this autocorrelation function (ACF) plot that **the series has a trend** (the ACF decreases slowly) and that **the series is seasonal and repeats every 12 months** (the ACF peaks every 12 months).

## Result of EDA

- The series **has a trend** (increasing).
- The series **is seasonal yearly** (the autocorrelation function peaks every 12 time units [months]).
- The series **shows a minor seasonal peak on June**.
- The series is **heteroskedastic** (the seasonal peaks have an increasing amplitude as time goes by).
