---
title: "Proyecto M1"
---

## About the series

<!-- TODO: Descripción completa del dataset. Pistas: (NSW household goods retail turnover), (millions of AUD), range (Apr 1982–Dec 2018), number of observations (441) -->

## Setup
<!-- PROTIP: en vez de picarle al botón de Preview, usar "quarto preview <filename> --no-browser", sin el --no-watch-inputs -->

```{r}
library(fpp3)
mi_serie <- aus_retail |> filter(`Series ID` == "A3349397X")
print(mi_serie)
tail(mi_serie)
```

## Plot of turnover

```{r}
mi_serie |> autoplot(Turnover) + labs(title="Household goods retail turnover in New South Wales, Australia", y="Sales turnover (millions of Australian dollars)")
```

We can notice three things about the series:

1. An upward trend.
2. A peak every year. Reading the data, it seems to be on December. It might have seasonality.
3. The size of the seasonal swings increases as time goes by. That means the series is heteroskedastic.

## Seasonal plot

```{r}
mi_serie |> gg_season(Turnover) + labs(title="Household goods retail turnover in New South Wales, Australia (turnover per year)", y="Sales turnover (millions of Australian dollars)")
```

This plot confirms not only that our series has yearly seasonality with a peak on December, it also reveals a minor peak on June (this could be due to accounting years in Australia customarily ending in June, which leads to seasonal purchases). <!-- TODO: poner bibliografía de esta hipótesis -->

## Subseries

```{r}
mi_serie |> gg_subseries(Turnover) + labs(title="Household goods retail turnover in New South Wales, Australia, grouped by month", y="Sales turnover (millions of Australian dollars)")
```

This plot shows that December is the month with the highest average. December is also the month that rises more steeply than the other months. That means the series is **heteroskedastic** (the seasonal effect scales with magnitude [multiplicative seasonality], making the series's variance non-constant).

## Autocorrelation

```{r}
mi_serie |> ACF(Turnover, lag_max=48) |> autoplot() + labs(title="Autocorrelation of household goods retail turnover in New South Wales, Australia", y="Value of autocorrelation function")
```

We can see in this autocorrelation function (ACF) plot that **the series has a trend** (the ACF decreases slowly) and that **the series is seasonal and repeats every 12 months** (the ACF peaks every 12 months).

## Result of EDA

- The series **has a trend** (increasing as shown from the time-domain plot and the ACF).
- The series **is seasonal yearly** (the autocorrelation function peaks every 12 time units [months], the subseries plot shows December as notoriously higher, the seasonal plot shows clear peaks in June and December).
- The series **shows a minor seasonal peak on June**.
- The series is **heteroskedastic** (the seasonal peaks have an increasing amplitude as time goes by, shown in the time-domain plot as well as the subseries plot).

## Box-Cox transform

We start by calculating our optimal lambda

```{r}
mi_serie |> features(Turnover, features=guerrero) |> pull(lambda_guerrero) -> lambda_optima
```

## Box-Cox transform

Since we calculated our lambda via Guerrero method, we can trust that this lambda will be the best one that stabilizes the growth trend and the amplitude of the seasonal peaks:

```{r}
mi_serie |> autoplot(box_cox(Turnover, lambda_optima)) + labs(title="Household goods retail turnover in New South Wales, Australia; scaled with Box-Cox transform, with lambda found via Guerrero method", y=paste("Sales turnover, scaled with lambda =", lambda_optima))
```

## Box-Cox transform

If we return to our vanilla series, we can notice how the vanilla series has a more exponential growth and how the amplitude of its seasonal peaks increases more as time goes by.

```{r}
mi_serie |> autoplot(Turnover) + labs(title="Household goods retail turnover in New South Wales, Australia", y="Sales turnover (millions of Australian dollars)")
```
